---
title: "JensenNarleskiSaraf_proj"
author: "Samantha Jensen, Jaclyn Narleski, Akanksha Saraf"
date: "2025-03-12"
output: html_document
---
Is there a significant difference between exposures for pollutants based on activity?
recreational, commercial, industrial, agricultural comparing specific pollutants (ANOVA)
comparing the NC air quality levels for exposure to the real levels that we are seeing.

next steps: 
- picking pollutants (2) (formaldehyde, diesel PM, maybe another)
- sorting columns into categories 
- find NC air quality standards for selected pollutants 




# Setup

```{r}

# Load packages
library(tidyverse)
library(here)

# Load data
exposure <- read.csv("ExposureConcentrationsRaw.csv")

```

# Data Wrangling & Tidying

```{r}

# Note that exposure concentration is in units ug/m3
# OR = On Road
# PT = Point source
# NR = Nonroad
# NP = Nonpoint

  
```

```{r}

# Seperating out potential VOCs that are considered carcinogens
carcinogens <- exposure %>%
  filter(Pollutant.Name %in% c("FORMALDEHYDE", "BENZENE", "TOULENE", 
                               "CARBON TETRACHLORIDE", "CARBON DISULFIDE", 
                               "1,3-BUTADIENE", "1,4-DICHLOROBENZENE"),
         FIPS == "37063")

carcinogens_summary <- carcinogens %>%
  group_by(Pollutant.Name) %>%
  summarise(average_exposure = mean(Total.Exposure.Conc, na.rm = TRUE)) %>%
  ungroup()

```

# Initial Figures 

```{r}

exposure_fig1 <- ggplot() +
  geom_jitter(data = carcinogens,
              aes(x = 1, y = Total.Exposure.Conc, color = "maroon2"),
              alpha = 0.7) +
  geom_point(data = carcinogens_summary,
             aes(x = 1, y = average_exposure), 
             size = 4,
             color = "darkred") +
  facet_wrap(~Pollutant.Name, scales = "free_y") +
  labs(title = "Air Exposure Concentrations of Carcinogenic VOCs in Durham, NC",
       y = "Total Exposure Concentration (mg/m3)",
       x = NULL,
       caption = "The above graphs display the varying exposure concentrations 
       of recorded volatilie organic compounds (VOCs) in the air during 2014 in 
       Durham, North Carolina. These specific VOCs are 
       known to potentially increase the risk of cancer, making them 
       carcinogens.") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.caption = element_text(hjust = 0.5),
        plot.title = element_text(hjust = 0.5, face = "bold", color = "darkred"),
        legend.position = "none")

exposure_fig1
```

```{r}
carcinogens_activity_summary <- carcinogens %>%
  group_by(Pollutant.Name) %>% 
  summarise(across(c(PT.StationaryPoint.Exposure.Conc:NP.AgricultureLivestock.Conc), mean))

carcinogens_long <- pivot_longer(carcinogens_activity_summary, cols = c(PT.StationaryPoint.Exposure.Conc:NP.AgricultureLivestock.Conc),  names_to = "Variable", values_to = "Value")
carcinogens_long <- carcinogens_long %>% 
  filter(Variable == "NR.CommercialLawnGarden.Exposure.Conc"| Variable == 
           "NR.ResidentialLawnGarden.Exposure.Conc"| Variable == 
           "NR.Construction.Exposure.Conc"| Variable ==
           "NR.CommercialEquipment.Exposure.Conc")
carcinogens_long <- carcinogens_long %>% 
  filter(Pollutant.Name == "1,3-BUTADIENE"| Pollutant.Name == "BENZENE"| Pollutant.Name =="FORMALDEHYDE")

carcinogens_activity_summary <- carcinogens_activity_summary %>% 
  filter(Pollutant.Name == "1,3-BUTADIENE"| Pollutant.Name == "BENZENE"| Pollutant.Name =="FORMALDEHYDE")


figure3 <- ggplot(carcinogens_long, aes(x = Pollutant.Name, y = Value)) + 
  geom_point() + 
  facet_wrap(~Variable) + 
  labs(x = "Pollutant Name", 
       y = "Exposure Concentration (mg/m3)", 
       title = "Non-Zero Carcinogen Exposure Levels", 
       caption = "This figure depicts the mean carcinogen exposure levels in Durham County from certain activities.")
figure3

```

