---
title: "JensenNarleskiSaraf_proj"
author: "Samantha Jensen, Jaclyn Narleski, Akanksha Saraf"
date: "2025-03-12"
output: html_document
---
Is there a significant difference between exposures for pollutants based on activity?
recreational, commercial, industrial, agricultural comparing specific pollutants (ANOVA)
comparing the NC air quality levels for exposure to the real levels that we are seeing.

next steps: 
- picking pollutants (2) (formaldehyde, diesel PM, maybe another)
- sorting columns into categories 
- find NC air quality standards for selected pollutants 

Unemployment Data: 
https://hdpulse.nimhd.nih.gov/data-portal/social/table?age=916&age_options=age25_1&demo=00012&demo_options=education_3&race=00&race_options=race_7&sex=0&sex_options=sexboth_1&socialtopic=090&socialtopic_options=social_6&statefips=37&statefips_options=area_states

Population Density: 
https://beautifydata.com/united-states-population/census/2020/population-density-by-county/north-carolina

Income: 
https://hdpulse.nimhd.nih.gov/data-portal/social/table?age=001&age_options=ageall_1&demo=00010&demo_options=income_3&race=00&race_options=race_7&sex=0&sex_options=sexboth_1&socialtopic=030&socialtopic_options=social_6&statefips=37&statefips_options=area_states
# Setup

```{r}

# Load packages
library(tidyverse)
library(here)
library(knitr)

# Load data
exposure <- read.csv("ExposureConcentrationsRaw.csv")
# Loading in dataset with urban/rural classification for each county in NC 
urban_rural <- read.xlsx("urban_rural.xlsx")
# dataset with unemployment and median family income 
census <- read.xlsx("census_combined.xlsx")


```
# Methods

## Data Source(s)

## Data Tidying

The raw dataset for exposure concentrations of air toxics in North Carolina contains the following information for each measurement of exposure: EPA region, county, FIPS code, FIPS tract, population, pollutant name, total exposure concentration, and several different sources that contributed to that total exposure concentration. The EPA divides these sources into different categories, as displayed below in Figure 1: 

Figure 1: National Air Toxics Assessment (NATA) Emission Sources

| On Road                       | Non-Road                           | Non-Point Stationary                       | Other                 |
|-------------------------------|------------------------------------|--------------------------------------------------------------------|
| Light Duty Gas On-Network     | Recreational & Pleasure Craft      | Industrial                                 | Fires                 |
| Light Duty Diesel On-Network  | Construction                       | Oil & Gas                                  | Biogenics             |
| Light Duty Gas Off-Network    | Commercial Lawn & Garden           | Commercial Cooking                         | Secondary Formation   |
| Light Duty Diesel Off-Network | Residential Lawn & Garden          | Solvents and Coatings                      | Background            |
| Heavy Duty Gas On-Network     | Agricultural Equipment             | Bulk Terminal Storage & Transfer (stage 1) |                       |                       
| Heavy Duty Diesel On-Network  | Commercial Equipment               | Fuel Combustion (except residential wood)  |                       |                       
| Heavy Duty Gas Off-Network    | Commercial Marine Vessel C1/C2     | Waste Disposal                             |                       |                       
| Heavy Duty Diesel Off-Network | Commercial Marine Vessel C3        | Agricultural Livestock                     |                       |                       
| Heavy Duty Diesel Hoteling    | Commercial Marine Vessels Underway | Miscellaneous Nonindustrial                |                       |                       
| Refueling                     | Locomotives                        |                                            |                       |                       
|                               | Airports (point source)            |                                            |                       |                     
|                               | Railroads (point source)           |                                            |                       |                       
|                               | All Other Non-Road Equipment       |                                            |                       |                       

"On Road" and "Non-Road" refer to mobile sources that emit air pollutants. "On Road" references mobile sources on the U.S. highway system that transport either passengers or freight. This includes passenger cars and light trucks (i.e. "light duty vehicles"); semi-trucks, RVs, delivery trucks, and buses (i.e. "heavy duty vehicles"); and each type of vehicle is separated by what type of fuel they use, gas or diesel. These types of vehicles are further categorized to "on-netork," or in motion, and "off-netork," or idle, as well as "refueling," or any vapor/spillage from fueling refills, or "hoteling," which is specific to heavy duty vehicles idling.

"Non-Road" references mobile sources that aren't on the U.S. highway system. This can range from recreational vehicles like snowcraft, boats, ATVs, dirt bikes to industrial/agricultural equipment and other forms of transportation like airports, railines ("railroads" and "locomotives"), and marine vessels. Marine vessels are categorized by engine size, with C1 and C2 considered smaller and C3 considered larger, as well as by "underway", or not anchored.

"Non-Point Stationary" refers to sources that aren't mobile or confined to single location. "Other" groups any other sources, including living organisms ("biogenics"), pollutants that form in the atmosphere from chemical reactions ("secondary"), and naturally-occurring emissions from human activity ("background").

# Data Wrangling & Tidying 

```{r}

# Select columns and rename for clarity
exposure_df <- exposure %>%
  select(County, Population, Pollutant.Name, Total.Exposure.Conc,
         OR.LightDuty.OffNetwork.Gas.Exposure.Conc,
         OR.LightDuty.OffNetwork.Diesel.Exposure.Conc,
         OR.HeavyDuty.OffNetwork.Gas.Exposure.Conc,
         OR.HeavyDuty.OffNetwork.Diesel.Exposure.Conc,
         OR.HeavyDuty.OnNetwork.Diesel.Exposure.Conc,
         OR.LightDuty.OnNetwork.Gas.Exposure.Conc,
         OR.LightDuty.OnNetwork.Diesel.Exposure.Conc,
         OR.HeavyDuty.OnNetwork.Gas.Exposure.Conc,
         OR.HeavyDuty.Hoteling.Exposure.Conc,
         OR.Refueling.Exposure.Conc,
         NR.Recreational.inc.PleasureCraft.Exposure.Conc,
         NR.Construction.Exposure.Conc,
         NR.Agriculture.Exposure.Conc,
         NR.CommercialEquipment.Exposure.Conc,
         NR.Locomotives.Exposure.Conc,
         NR.Point.Airports.Exposure.Conc,
         NR..Point.Railyards.Exposure.Conc) %>%
  rename("OR.Light.Off.Gas" = OR.LightDuty.OffNetwork.Gas.Exposure.Conc,
         "OR.Light.Off.Diesel" = OR.LightDuty.OffNetwork.Diesel.Exposure.Conc,
         "OR.Heavy.Off.Gas" = OR.HeavyDuty.OffNetwork.Gas.Exposure.Conc,
         "OR.Heavy.Off.Diesel" = OR.HeavyDuty.OffNetwork.Diesel.Exposure.Conc,
         "OR.Light.On.Gas" = OR.LightDuty.OnNetwork.Gas.Exposure.Conc,
         "OR.Light.On.Diesel" = OR.LightDuty.OnNetwork.Diesel.Exposure.Conc,
         "OR.Heavy.On.Gas" = OR.HeavyDuty.OnNetwork.Gas.Exposure.Conc,
         "OR.Heavy.On.Diesel" = OR.HeavyDuty.OnNetwork.Diesel.Exposure.Conc,
         "OR.Heavy.Idling" = OR.HeavyDuty.Hoteling.Exposure.Conc,
         "OR.Refueling" = OR.Refueling.Exposure.Conc,
         "NR.Recreational.Vehicles" = 
           NR.Recreational.inc.PleasureCraft.Exposure.Conc,
         "NR.Construction" = NR.Construction.Exposure.Conc,
         "NR.Agriculture" = NR.Agriculture.Exposure.Conc,
         "NR.Commerical.Equip" = NR.CommercialEquipment.Exposure.Conc,
         "NR.Locomotives" = NR.Locomotives.Exposure.Conc,
         "NR.Airpoints" = NR.Point.Airports.Exposure.Conc,
         "NR.Railyards" = NR..Point.Railyards.Exposure.Conc)

```

```{r}
# making rural/urban data into a factor 
urban_rural$URBAN_RURAL <- as.factor(urban_rural$URBAN_RURAL)
# filtering census data 
census_select <- census %>% 
  select(County, 
         `Median Family Income (Dollars)`:`Population Density 
         (Persons Per SqMile)`)
# making income a numeric class 
census_select$`Median Family Income (Dollars)` <- as.numeric(gsub("[^0-9]", "", 
                                census_select$`Median Family Income (Dollars)`))
census_select$`Population Density (Persons Per SqMile)` <-  as.numeric(
  gsub("[^0-9,.]", "", census_select$`Population Density (Persons Per SqMile)`))
```

```{r}
# Making datasets for the Pollutants of interest 
formaldehyde <- exposure_df %>% 
  filter(Pollutant.Name == "FORMALDEHYDE") %>% 
  group_by(County) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  ungroup()
# subtracting entire state
formaldehyde <- formaldehyde[-34,]
# adding columns that sum all of the nonroad and onroad exposures for each county 
# formaldehyde dataframe still has all of the exposure activities so it can be 
# used for other analysis if we want 
formaldehyde <- formaldehyde %>% 
  mutate(Onroad_total = select(., OR.Light.Off.Gas:OR.Refueling) %>% 
      rowSums(na.rm = TRUE)) %>% 
  mutate(Nonroad_total = select(., NR.Recreational.Vehicles:NR.Railyards) %>% 
      rowSums(na.rm = TRUE)) 

# joining urban/rural column 
formaldehyde_select <- left_join(formaldehyde, urban_rural, by = "County")
# selecting necessary columns 
formaldehyde_select <- formaldehyde_select %>% 
  select(County, Population, Total.Exposure.Conc, Onroad_total:URBAN_RURAL) 
# joining with census data 
formaldehyde_complete <- left_join(formaldehyde_select, census_select, by = "County")
# making county into factor
formaldehyde_complete$County <- as.factor(formaldehyde_complete$County)

```

```{r}

## Benzene 
benzene <- exposure_df %>% 
  filter(Pollutant.Name == "BENZENE") %>% 
  group_by(County) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  ungroup()
# subtracting entire state
benzene <- benzene[-34,]
# adding columns that sum all of the nonroad and onroad exposures for each county 
benzene <- benzene %>% 
  mutate(Onroad_total = select(., OR.Light.Off.Gas:OR.Refueling) %>% 
           rowSums(na.rm = TRUE)) %>% 
  mutate(Nonroad_total = select(., NR.Recreational.Vehicles:NR.Railyards) %>% 
           rowSums(na.rm = TRUE)) 

# joining urban/rural column 
benzene_select <- left_join(benzene, urban_rural, by = "County")
# selecting necessary columns 
benzene_select <- benzene_select %>% 
  select(County, Population, Total.Exposure.Conc, Onroad_total:URBAN_RURAL) 
# joining with census data 
benzene_complete <- left_join(benzene_select, census_select, by = "County") 
# making county into factor 
benzene_complete$County <- as.factor(benzene_complete$County)

```

```{r}

## Diesel 
diesel <- exposure_df %>% 
  filter(Pollutant.Name == "DIESEL PM") %>% 
  group_by(County) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  ungroup()
# subtracting entire state
diesel <- diesel[-34,]
# adding columns that sum all of the nonroad and onroad exposures for each county 
diesel <- diesel %>% 
  mutate(
    Onroad_total = 
      select(., OR.Light.Off.Gas:OR.Refueling) %>% 
      rowSums(na.rm = TRUE)) %>% 
  mutate(Nonroad_total = select(., NR.Recreational.Vehicles:NR.Railyards) %>% 
           rowSums(na.rm = TRUE)) 

# joining urban/rural column 
diesel_select <- left_join(diesel, urban_rural, by = "County")
# selecting necessary columns 
diesel_select <- diesel_select %>% 
  select(County, Population, Total.Exposure.Conc, Onroad_total:URBAN_RURAL)
  
# joining with census data 
diesel_complete <- left_join(diesel_select, census_select, by = "County") 
diesel_complete$County <- as.factor(diesel_complete$County)

## Can still join the cancer data in if we want to 

```

## Exploratory Figures: 

## Analysis: 
### Formaldehyde Model 
```{r}
# Visualizing the relationship between response and descriptor variables 
(scatter_population <- ggplot(formaldehyde_complete, aes(x = Population, y = Total.Exposure.Conc)) + 
  geom_point())
(scatter_income <- ggplot(formaldehyde_complete, aes(x = `Median Family Income (Dollars)`, y = Total.Exposure.Conc)) + 
    geom_point())
(scatter_urban <- ggplot(formaldehyde_complete, aes(x = URBAN_RURAL, y = Total.Exposure.Conc)) + 
    geom_point())
(scatter_density <- ggplot(formaldehyde_complete, aes(x = `Population Density (Persons Per SqMile)`, y = Total.Exposure.Conc)) +
    geom_point())
(scatter_unemployment <- ggplot(formaldehyde_complete, aes(x =`Unemployment(Percent)`, y = Total.Exposure.Conc)) + 
    geom_point())


## workflow for models 
# 1. Colinearity (cannot include the variables that have a pearsons R greater than 0.6)
# 2. creating models 
# 3. interpreting results 

```

### Benzene Model 
```{r}

```

# Diesel Model 
```{r}

```

