---
title: "JensenNarleskiSaraf_proj"
author: "Samantha Jensen, Jaclyn Narleski, Akanksha Saraf"
date: "2025-03-12"
output:
  pdf_document: default
  html_document: default
---
Is there a significant difference between exposures for pollutants based on activity?
recreational, commercial, industrial, agricultural comparing specific pollutants (ANOVA)
comparing the NC air quality levels for exposure to the real levels that we are seeing.

next steps: 
- picking pollutants (2) (formaldehyde, diesel PM, maybe another)
- sorting columns into categories 
- find NC air quality standards for selected pollutants 

Unemployment Data: 
https://hdpulse.nimhd.nih.gov/data-portal/social/table?age=916&age_options=age25_1&demo=00012&demo_options=education_3&race=00&race_options=race_7&sex=0&sex_options=sexboth_1&socialtopic=090&socialtopic_options=social_6&statefips=37&statefips_options=area_states

Population Density: 
https://beautifydata.com/united-states-population/census/2020/population-density-by-county/north-carolina

Income: 
https://hdpulse.nimhd.nih.gov/data-portal/social/table?age=001&age_options=ageall_1&demo=00010&demo_options=income_3&race=00&race_options=race_7&sex=0&sex_options=sexboth_1&socialtopic=030&socialtopic_options=social_6&statefips=37&statefips_options=area_states

# Setup

```{r}

# Load packages
library(tidyverse)
library(here)
library(knitr)
library(readxl)
library(patchwork)

# Load data
exposure <- read.csv("ExposureConcentrationsRaw.csv")
# Loading in dataset with urban/rural classification for each county in NC 
urban_rural <- read_excel("urban_rural.xlsx")
# dataset with unemployment and median family income 
census <- read_excel("census_combined.xlsx")


```

Figure 1: National Air Toxics Assessment (NATA) Emission Sources

| On Road                       | Non-Road                           | Non-Point Stationary                       | Other                 |
|-------------------------------|------------------------------------|--------------------------------------------------------------------|
| Light Duty Gas On-Network     | Recreational & Pleasure Craft      | Industrial                                 | Fires                 |
| Light Duty Diesel On-Network  | Construction                       | Oil & Gas                                  | Biogenics             |
| Light Duty Gas Off-Network    | Commercial Lawn & Garden           | Commercial Cooking                         | Secondary Formation   |
| Light Duty Diesel Off-Network | Residential Lawn & Garden          | Solvents and Coatings                      | Background            |
| Heavy Duty Gas On-Network     | Agricultural Equipment             | Bulk Terminal Storage & Transfer (stage 1) |                       |                       
| Heavy Duty Diesel On-Network  | Commercial Equipment               | Fuel Combustion (except residential wood)  |                       |                       
| Heavy Duty Gas Off-Network    | Commercial Marine Vessel C1/C2     | Waste Disposal                             |                       |                       
| Heavy Duty Diesel Off-Network | Commercial Marine Vessel C3        | Agricultural Livestock                     |                       |                       
| Heavy Duty Diesel Hoteling    | Commercial Marine Vessels Underway | Miscellaneous Nonindustrial                |                       |                       
| Refueling                     | Locomotives                        |                                            |                       |                       
|                               | Airports (point source)            |                                            |                       |                     
|                               | Railroads (point source)           |                                            |                       |                       
|                               | All Other Non-Road Equipment       |                                            |                       |                       

## Data Tidying & Wrangling Code

#### Exposure Data

```{r}

# Select columns and rename for clarity
exposure_df <- exposure %>%
  select(County, Population, Pollutant.Name, Total.Exposure.Conc,
         OR.LightDuty.OffNetwork.Gas.Exposure.Conc,
         OR.LightDuty.OffNetwork.Diesel.Exposure.Conc,
         OR.HeavyDuty.OffNetwork.Gas.Exposure.Conc,
         OR.HeavyDuty.OffNetwork.Diesel.Exposure.Conc,
         OR.HeavyDuty.OnNetwork.Diesel.Exposure.Conc,
         OR.LightDuty.OnNetwork.Gas.Exposure.Conc,
         OR.LightDuty.OnNetwork.Diesel.Exposure.Conc,
         OR.HeavyDuty.OnNetwork.Gas.Exposure.Conc,
         OR.HeavyDuty.Hoteling.Exposure.Conc,
         OR.Refueling.Exposure.Conc,
         NR.Recreational.inc.PleasureCraft.Exposure.Conc,
         NR.Construction.Exposure.Conc,
         NR.Agriculture.Exposure.Conc,
         NR.CommercialEquipment.Exposure.Conc,
         NR.Locomotives.Exposure.Conc,
         NR.Point.Airports.Exposure.Conc,
         NR..Point.Railyards.Exposure.Conc) %>%
  rename("OR.Light.Off.Gas" = OR.LightDuty.OffNetwork.Gas.Exposure.Conc,
         "OR.Light.Off.Diesel" = OR.LightDuty.OffNetwork.Diesel.Exposure.Conc,
         "OR.Heavy.Off.Gas" = OR.HeavyDuty.OffNetwork.Gas.Exposure.Conc,
         "OR.Heavy.Off.Diesel" = OR.HeavyDuty.OffNetwork.Diesel.Exposure.Conc,
         "OR.Light.On.Gas" = OR.LightDuty.OnNetwork.Gas.Exposure.Conc,
         "OR.Light.On.Diesel" = OR.LightDuty.OnNetwork.Diesel.Exposure.Conc,
         "OR.Heavy.On.Gas" = OR.HeavyDuty.OnNetwork.Gas.Exposure.Conc,
         "OR.Heavy.On.Diesel" = OR.HeavyDuty.OnNetwork.Diesel.Exposure.Conc,
         "OR.Heavy.Idling" = OR.HeavyDuty.Hoteling.Exposure.Conc,
         "OR.Refueling" = OR.Refueling.Exposure.Conc,
         "NR.Recreational.Vehicles" = 
           NR.Recreational.inc.PleasureCraft.Exposure.Conc,
         "NR.Construction" = NR.Construction.Exposure.Conc,
         "NR.Agriculture" = NR.Agriculture.Exposure.Conc,
         "NR.Commerical.Equip" = NR.CommercialEquipment.Exposure.Conc,
         "NR.Locomotives" = NR.Locomotives.Exposure.Conc,
         "NR.Airpoints" = NR.Point.Airports.Exposure.Conc,
         "NR.Railyards" = NR..Point.Railyards.Exposure.Conc) %>%
  mutate(County = as.factor(County))

```

#### Urban/Rural & Census Data 

```{r}
# Making rural/urban data into a factor 
urban_rural$URBAN_RURAL <- as.factor(urban_rural$URBAN_RURAL)
# Filtering census data 
census_select <- census %>% 
  select(County, 
         `Median Family Income (Dollars)`:
           `Population Density (Persons Per SqMile)`)
# Making income a numeric class 
census_select$`Median Family Income (Dollars)` <- as.numeric(gsub("[^0-9]", "", 
                                census_select$`Median Family Income (Dollars)`))
census_select$`Population Density (Persons Per SqMile)` <-  as.numeric(gsub("[^0-9.]", "", census_select$`Population Density (Persons Per SqMile)`))
```

#### Formaldehyde Data

```{r}

formaldehyde <- exposure_df %>% 
  filter(Pollutant.Name == "FORMALDEHYDE") %>% 
  group_by(County) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  ungroup()
# subtracting entire state
formaldehyde <- formaldehyde[-34,]
# adding columns that sum all of the nonroad and onroad exposures for each county 
# formaldehyde dataframe still has all of the exposure activities so it can be 
# used for other analysis if we want 
formaldehyde <- formaldehyde %>% 
  mutate(Onroad_total = select(., OR.Light.Off.Gas:OR.Refueling) %>% 
      rowSums(na.rm = TRUE)) %>% 
  mutate(Nonroad_total = select(., NR.Recreational.Vehicles:NR.Railyards) %>% 
      rowSums(na.rm = TRUE)) 

# joining urban/rural column 
formaldehyde_select <- left_join(formaldehyde, urban_rural, by = "County")
# selecting necessary columns 
formaldehyde_select <- formaldehyde_select %>% 
  select(County, Population, Total.Exposure.Conc, Onroad_total:URBAN_RURAL) 
# joining with census data 
formaldehyde_complete <- left_join(formaldehyde_select, 
                                   census_select, by = "County")
# making county into factor
formaldehyde_complete$County <- as.factor(formaldehyde_complete$County)

```

#### Benzene Data

```{r}

benzene <- exposure_df %>% 
  filter(Pollutant.Name == "BENZENE") %>% 
  group_by(County) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  ungroup()
# subtracting entire state
benzene <- benzene[-34,]
# adding columns that sum all of the nonroad and onroad exposures for each county 
benzene <- benzene %>% 
  mutate(Onroad_total = select(., OR.Light.Off.Gas:OR.Refueling) %>% 
           rowSums(na.rm = TRUE)) %>% 
  mutate(Nonroad_total = select(., NR.Recreational.Vehicles:NR.Railyards) %>% 
           rowSums(na.rm = TRUE)) 

# joining urban/rural column 
benzene_select <- left_join(benzene, urban_rural, by = "County")
# selecting necessary columns 
benzene_select <- benzene_select %>% 
  select(County, Population, Total.Exposure.Conc, Onroad_total:URBAN_RURAL) 
# joining with census data 
benzene_complete <- left_join(benzene_select, census_select, by = "County") 
# making county into factor 
benzene_complete$County <- as.factor(benzene_complete$County)

```

#### Diesel Data

```{r}

## Diesel 
diesel <- exposure_df %>% 
  filter(Pollutant.Name == "DIESEL PM") %>% 
  group_by(County) %>% 
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  ungroup()
# subtracting entire state
diesel <- diesel[-34,]
# adding columns that sum all of the nonroad and onroad exposures for each county 
diesel <- diesel %>% 
  mutate(
    Onroad_total = 
      select(., OR.Light.Off.Gas:OR.Refueling) %>% 
      rowSums(na.rm = TRUE)) %>% 
  mutate(Nonroad_total = select(., NR.Recreational.Vehicles:NR.Railyards) %>% 
           rowSums(na.rm = TRUE)) 

# joining urban/rural column 
diesel_select <- left_join(diesel, urban_rural, by = "County")
# selecting necessary columns 
diesel_select <- diesel_select %>% 
  select(County, Population, Total.Exposure.Conc, Onroad_total:URBAN_RURAL)
  
# joining with census data 
diesel_complete <- left_join(diesel_select, census_select, by = "County") 
diesel_complete$County <- as.factor(diesel_complete$County)

## Can still join the cancer data in if we want to 

```

# Analysis

#### Summary Statistics 

```{r}
# descriptive statistics for results section: 

mean_form <- mean(formaldehyde_complete$Total.Exposure.Conc)
sd_form <- sd(formaldehyde_complete$Total.Exposure.Conc)
mean_b <- mean(benzene_complete$Total.Exposure.Conc)
sd_b <- sd(benzene_complete$Total.Exposure.Conc)
mean_D <- mean(diesel_complete$Total.Exposure.Conc)
sd_d <- sd(diesel_complete$Total.Exposure.Conc)

mean_population <- mean(formaldehyde_complete$Population)
sd_pop <- sd(formaldehyde_complete$Population)
mean_population_density <- 
  mean(formaldehyde_complete$`Population Density (Persons Per SqMile)`)
sd_density <- 
  sd(formaldehyde_complete$`Population Density (Persons Per SqMile)`)
mean_income <- mean(formaldehyde_complete$`Median Family Income (Dollars)`)
sd_income <- sd(formaldehyde_complete$`Median Family Income (Dollars)`)
mean_unemploy <- mean(formaldehyde_complete$`Unemployment(Percent)`)
sd_unemploy <- sd(formaldehyde_complete$`Unemployment(Percent)`)

```

#### Onroad & Nonroad Data

```{r}

formaldehyde_onroad <- formaldehyde_complete %>%
  ggplot(aes(x = Onroad_total)) +
  geom_histogram(bins = 60, fill = "skyblue2", color = "slateblue4") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "slateblue4"),
        axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
  labs(title = "Onroad Exposure",
       x = "Exposure Concentration (ug/m3)",
       y = "Count")

formaldehyde_nonroad <- formaldehyde_complete %>%
  ggplot(aes(x = Nonroad_total)) +
  geom_histogram(bins = 60, fill = "skyblue2", color = "slateblue4") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "slateblue4"),
        axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
  labs(title = "Nonroad Exposure",
       x = "Exposure Concentration (ug/m3)",
       y = "Count")

form_plot <- (formaldehyde_onroad + formaldehyde_nonroad) + 
  plot_annotation(title = "Formaldehyde Onroad & Nonroad Concentrations ",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, face = "bold", 
                                              color = "slateblue4")))
plot(form_plot)

benzene_onroad <- benzene_complete %>%
  ggplot(aes(x = Onroad_total)) +
  geom_histogram(bins = 60, fill = "palegreen2", color = "darkgreen") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
        axis.title.x = element_text(face = "bold", color = "darkgreen"),
        axis.title.y = element_text(face = "bold", color = "darkgreen"),
        panel.background = element_rect(fill = "azure"),
        panel.grid.major = element_line(color = "azure3"),
        panel.grid.minor = element_line(color = "azure2")) +
  labs(title = "Onroad Exposure",
       x = "Exposure Concentration (ug/m3)",
       y = "Count")

benzene_nonroad <- benzene_complete %>%
  ggplot(aes(x = Nonroad_total)) +
  geom_histogram(bins = 60, fill = "palegreen2", color = "darkgreen") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
        axis.title.x = element_text(face = "bold", color = "darkgreen"),
        axis.title.y = element_text(face = "bold", color = "darkgreen"),
        panel.background = element_rect(fill = "azure"),
        panel.grid.major = element_line(color = "azure3"),
        panel.grid.minor = element_line(color = "azure2")) +
  labs(title = "Nonroad Exposure",
       x = "Exposure Concentration (ug/m3)",
       y = "Count")

ben_plot <- (benzene_onroad + benzene_nonroad) + 
  plot_annotation(title = "Benzene Onroad & Nonroad Concentrations ",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, face = "bold", 
                                              color = "darkgreen")))
plot(ben_plot)

diesel_onroad <- diesel_complete %>%
  ggplot(aes(x = Onroad_total)) +
  geom_histogram(bins = 60, fill = "palevioletred", color = "violetred4") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) +
  labs(title = "Onroad Exposure",
       x = "Exposure Concentration (ug/m3)",
       y = "Count")

diesel_nonroad <- diesel_complete %>%
  ggplot(aes(x = Nonroad_total)) +
  geom_histogram(bins = 60, fill = "palevioletred", color = "violetred4") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) +
  labs(title = "Nonroad Exposure",
       x = "Exposure Concentration (ug/m3)",
       y = "Count")

diesel_plot <- (diesel_onroad + diesel_nonroad) + 
  plot_annotation(title = "Diesel PM Onroad & Nonroad Concentrations ",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, face = "bold", 
                                              color = "violetred4")))
plot(diesel_plot)

```

#### Formaldehyde Model 

```{r}

# Visualizing the relationship between response and descriptor variables 
(scatter_population <- ggplot(formaldehyde_complete, 
                              aes(x = Population, y = Total.Exposure.Conc)) + 
  geom_point(color = "skyblue3") + 
   theme(axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
   labs(x = "Population", 
        y = "ug/m3"))

(scatter_income <- ggplot(formaldehyde_complete, 
                          aes(x = `Median Family Income (Dollars)`, 
                              y = Total.Exposure.Conc)) + 
    geom_point(color = "skyblue3") + 
    scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
    theme(axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
   labs(x = "Median Family Income ($)", 
        y = "ug/m3"))

(scatter_urban <- ggplot(formaldehyde_complete, 
                         aes(x = URBAN_RURAL, y = Total.Exposure.Conc)) + 
    geom_point(color = "skyblue3") + 
    theme(axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
   labs(x = "Urban or Rural County", 
        y = "ug/m3"))

(scatter_density <- ggplot(formaldehyde_complete, 
                           aes(x = `Population Density (Persons Per SqMile)`, 
                               y = Total.Exposure.Conc)) +
    geom_point(color = "skyblue3") + 
    theme(axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
   labs(x = "Population Density (Person/sq mile)", 
        y = "ug/m3"))

(scatter_unemployment <- ggplot(formaldehyde_complete, 
                                aes(x =`Unemployment(Percent)`, 
                                    y = Total.Exposure.Conc)) + 
    geom_point(color = "skyblue3") + 
    theme(axis.title.x = element_text(face = "bold", color = "slateblue4"),
        axis.title.y = element_text(face = "bold", color = "slateblue4"),
        panel.background = element_rect(fill = "aliceblue"),
        panel.grid.major = element_line(color = "lavenderblush3"),
        panel.grid.minor = element_line(color = "lavenderblush2")) +
   labs(x = "Unemployment (%)", 
        y = "ug/m3"))

final_plot <- ((scatter_density + scatter_urban) / 
  (scatter_population + scatter_unemployment) / scatter_income) + 
  plot_annotation(title = "Formaldehyde Concentrations vs. County Demographics",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, face = "bold", 
                                              color = "slateblue4")))
plot(final_plot)

# Perform correlation tests to test for collinearity

cor.test(formaldehyde_complete$Population, 
         formaldehyde_complete$`Median Family Income (Dollars)`) 
# p < 0.01, cor = 0.4802

cor.test(formaldehyde_complete$Population, 
         formaldehyde_complete$`Population Density (Persons Per SqMile)`) 
# p < 0.01, cor = 0.9144

cor.test(formaldehyde_complete$`Unemployment(Percent)`, 
         formaldehyde_complete$`Median Family Income (Dollars)`)
# p < 0.01, cor = -0.4911

cor.test(formaldehyde_complete$`Median Family Income (Dollars)`,
         formaldehyde_complete$`Population Density (Persons Per SqMile)`)
# p < 0.01, cor = 0.402

# Generate regression model(s)
F_model1 <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
               formaldehyde_complete)
summary(F_model1)
plot(F_model1)

F_model2 <- lm(Total.Exposure.Conc ~ `Population Density (Persons Per SqMile)` + 
                 URBAN_RURAL, formaldehyde_complete)
summary(F_model2)
plot(F_model2)

formaldehyde_nooutlier <- formaldehyde_complete[-92,]
formaldehyde_nooutlier <- formaldehyde_nooutlier[-60,]
model1_nooutlier <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
                       formaldehyde_nooutlier)
summary(model1_nooutlier)
plot(model1_nooutlier)

model2_nooutlier <- lm(Total.Exposure.Conc ~ 
                         `Population Density (Persons Per SqMile)` + 
                         URBAN_RURAL, formaldehyde_nooutlier)
summary(model2_nooutlier)
plot(model2_nooutlier)

model3 <- lm(Total.Exposure.Conc ~ URBAN_RURAL + 
               `Median Family Income (Dollars)` + 
               Population, formaldehyde_nooutlier)
summary(model3)
plot(model3)

AIC1 <- AIC(model1_nooutlier)
AIC2 <- AIC(model2_nooutlier)
AIC3 <- AIC(model3)

# Model 1 without outliers has the lowest AIC value and is the one that I will write the results about. 

(histogram_form <- ggplot(formaldehyde_complete, aes(x = Total.Exposure.Conc)) + 
  geom_histogram())

formaldehyde_complete <- formaldehyde_complete %>% 
  mutate(log_concentrations = log10(Total.Exposure.Conc))

(hist_log_F <- ggplot(formaldehyde_complete, aes(x = log_concentrations)) + 
  geom_histogram())

F_model_log1 <- lm(log_concentrations ~ Population + URBAN_RURAL,
                 formaldehyde_complete)
summary(F_model_log1)
plot(F_model_log1)

f_log_nooutlier <- formaldehyde_complete[-c(60,92),]

F_model_log1_noout <- lm(log_concentrations ~ Population + URBAN_RURAL,
                 f_log_nooutlier)
summary(F_model_log1_noout)
plot(F_model_log1_noout)

f_log_nout2 <- f_log_nooutlier[-c(32,41),]
f_log_noout_model2 <- lm(log_concentrations ~ Population + URBAN_RURAL,
                 f_log_nout2)
summary(f_log_noout_model2)
plot(f_log_noout_model2)

f_final_outlier <- f_log_nout2[-c(26,33),]
log_model3 <- lm(log_concentrations ~ Population + URBAN_RURAL,
                 f_final_outlier)
summary(log_model3)
plot(log_model3)

log_model4 <- lm(log_concentrations ~ `Population Density (Persons Per SqMile)` + URBAN_RURAL,
                 f_final_outlier)
summary(log_model4)
plot(log_model4)

log_model5 <- lm(log_concentrations ~ URBAN_RURAL + 
               `Median Family Income (Dollars)` + 
               Population, f_final_outlier)
summary(log_model5)
plot(log_model5)

log_model6 <- lm(log_concentrations ~ URBAN_RURAL + Population + `Unemployment(Percent)`, f_final_outlier)
summary(log_model6)
plot(log_model6)

AIC3 <- AIC(log_model3)
AIC4 <- AIC(log_model4)
AIC5 <- AIC(log_model5)
AIC6 <- AIC(log_model6)
```

#### Benzene Model 

```{r}

(scatter_populationB <- ggplot(benzene_complete, 
                              aes(x = Population, y = Total.Exposure.Conc)) + 
  geom_point(color = "palegreen4") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
        axis.title.x = element_text(face = "bold", color = "darkgreen"),
        axis.title.y = element_text(face = "bold", color = "darkgreen"),
        panel.background = element_rect(fill = "azure"),
        panel.grid.major = element_line(color = "azure3"),
        panel.grid.minor = element_line(color = "azure2")) +
  labs(x = "Population",
       y = "ug/m3"))

(scatter_incomeB <- ggplot(benzene_complete, 
                          aes(x = `Median Family Income (Dollars)`, 
                              y = Total.Exposure.Conc)) + 
  geom_point(color = "palegreen4") +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
        axis.title.x = element_text(face = "bold", color = "darkgreen"),
        axis.title.y = element_text(face = "bold", color = "darkgreen"),
        panel.background = element_rect(fill = "azure"),
        panel.grid.major = element_line(color = "azure3"),
        panel.grid.minor = element_line(color = "azure2")) +
  labs(x = "Median Family Income ($)",
       y = "ug/m3"))

(scatter_urbanB <- ggplot(benzene_complete, 
                         aes(x = URBAN_RURAL, y = Total.Exposure.Conc)) + 
    geom_point(color = "palegreen4") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
          axis.title.x = element_text(face = "bold", color = "darkgreen"),
          axis.title.y = element_text(face = "bold", color = "darkgreen"),
          panel.background = element_rect(fill = "azure"),
          panel.grid.major = element_line(color = "azure3"),
          panel.grid.minor = element_line(color = "azure2")) +
  labs(x = "Urban or Rural County",
       y = "ug/m3"))

(scatter_densityB <- ggplot(benzene_complete, 
                           aes(x = `Population Density (Persons Per SqMile)`, 
                               y = Total.Exposure.Conc)) +
    geom_point(color = "palegreen4") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
          axis.title.x = element_text(face = "bold", color = "darkgreen"),
          axis.title.y = element_text(face = "bold", color = "darkgreen"),
          panel.background = element_rect(fill = "azure"),
          panel.grid.major = element_line(color = "azure3"),
          panel.grid.minor = element_line(color = "azure2")) +
  labs(x = "Population Density (Persons/sq Mile)",
       y = "ug/m3"))


(scatter_unemploymentB <- ggplot(benzene_complete, 
                                aes(x =`Unemployment(Percent)`, 
                                    y = Total.Exposure.Conc)) + 
    geom_point(color = "palegreen4") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "darkgreen"),
          axis.title.x = element_text(face = "bold", color = "darkgreen"),
          axis.title.y = element_text(face = "bold", color = "darkgreen"),
          panel.background = element_rect(fill = "azure"),
          panel.grid.major = element_line(color = "azure3"),
          panel.grid.minor = element_line(color = "azure2")) +
  labs(x = "Unemployment (%)",
       y = "ug/m3"))


final_plotB <- ((scatter_densityB + scatter_urbanB) / 
  (scatter_populationB + scatter_unemploymentB) / scatter_incomeB) + 
  plot_annotation(title = "Benzene Concentrations vs. County Demographics",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, face = "bold", 
                                              color = "darkgreen")))
plot(final_plotB)

# Perform correlation tests to test for collinearity
cor.test(benzene_complete$Population, 
         benzene_complete$`Median Family Income (Dollars)`) 
# p < 0.01, cor = 0.4802

cor.test(benzene_complete$Population, 
         benzene_complete$`Population Density (Persons Per SqMile)`) 
# p < 0.01, cor = 0.9144

cor.test(benzene_complete$`Unemployment(Percent)`, 
         benzene_complete$`Median Family Income (Dollars)`)
# p < 0.01, cor = -0.4911

cor.test(benzene_complete$`Median Family Income (Dollars)`,
         benzene_complete$`Population Density (Persons Per SqMile)`)
# p < 0.01, cor = 0.402

# Generate regression models
B_model1 <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
               benzene_complete)
summary(B_model1)
plot(B_model1)
AIC(B_model1) # = 519.3985

benzene_nooutlier <- benzene_complete[-60,]
benzene_nooutlier <- benzene_nooutlier[-92,]

Bmodel1_nooutlier <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
                       benzene_nooutlier)
summary(Bmodel1_nooutlier)
plot(Bmodel1_nooutlier)
AIC(Bmodel1_nooutlier) # = 447.2516

benzene_nooutlier <- benzene_nooutlier[-91,]

B_model2 <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
               benzene_nooutlier)
summary(B_model2)
plot(B_model2)
AIC(B_model2) # = 441.7628

benzene_complete <- benzene_complete %>%
  mutate(log_conc = log10(Total.Exposure.Conc))

B_model_log <- lm(log_conc ~ Population + URBAN_RURAL, 
               benzene_complete)
summary(B_model_log)
plot(B_model_log)
AIC(B_model_log) #32.28993

B_model_log2 <- lm(log_conc ~ `Unemployment(Percent)` + URBAN_RURAL, 
               benzene_complete)

summary(B_model_log2)
plot(B_model_log2)
AIC(B_model_log2)

log_nooutliers <- benzene_complete[-c(60,48,37),]

B_model_log3 <- lm(log_conc ~ `Unemployment(Percent)` + URBAN_RURAL, 
               log_nooutliers)

summary(B_model_log3)
plot(B_model_log3)
AIC(B_model_log3)

B_model_log4 <- lm(log_conc ~ `Median Family Income (Dollars)` + URBAN_RURAL, 
               log_nooutliers)

summary(B_model_log4)
plot(B_model_log4)
AIC(B_model_log4)

```

# Diesel Model 
```{r}

(scatter_populationD <- ggplot(diesel_complete, 
                              aes(x = Population, y = Total.Exposure.Conc)) + 
  geom_point(color = "palevioletred") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) +
    labs(x = "Population", 
        y = "ug/m3"))

(scatter_incomeD <- ggplot(diesel_complete, 
                          aes(x = `Median Family Income (Dollars)`, 
                              y = Total.Exposure.Conc)) + 
  geom_point(color = "palevioletred") +
     scale_x_continuous(labels = function(x) format(x, scientific = FALSE)) +
   theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) + 
     labs(x = "Median Family Income ($)", 
        y = "ug/m3"))

(scatter_urbanD <- ggplot(diesel_complete, 
                         aes(x = URBAN_RURAL, y = Total.Exposure.Conc)) + 
   geom_point(color = "palevioletred") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) + 
     labs(x = "Urban or Rural County", 
        y = "ug/m3"))

(scatter_densityD <- ggplot(diesel_complete, 
                           aes(x = `Population Density (Persons Per SqMile)`, 
                               y = Total.Exposure.Conc)) +
    geom_point(color = "palevioletred") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) + 
     labs(x = "Population Density (Persons/sq Mile)", 
        y = "ug/m3"))

(scatter_unemploymentD <- ggplot(diesel_complete, 
                                aes(x =`Unemployment(Percent)`, 
                                    y = Total.Exposure.Conc)) + 
    geom_point(color = "palevioletred") +
   theme(plot.title = element_text(hjust = 0.5, face = "bold", 
                                  color = "violetred4"),
        axis.title.x = element_text(face = "bold", color = "violetred4"),
        axis.title.y = element_text(face = "bold", color = "violetred4"),
        panel.background = element_rect(fill = "seashell"),
        panel.grid.major = element_line(color = "mistyrose3"),
        panel.grid.minor = element_line(color = "mistyrose2")) +
     labs(x = "Unemployment (%)", 
        y = "ug/m3")) 


final_plot_diesel <- ((scatter_densityD + scatter_urbanD) / 
  (scatter_populationD + scatter_unemploymentD) / scatter_incomeD) + 
  plot_annotation(title = "Diesel Concentrations vs. County Demographics",
                  theme = theme(
                    plot.title = element_text(hjust = 0.5, face = "bold", 
                                              color = "violetred4")))
plot(final_plot_diesel)

# Perform correlation tests to test for collinearity

cor.test(diesel_complete$Population, 
         diesel_complete$`Median Family Income (Dollars)`) 

# Generate regression models
D_model1 <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
               diesel_complete)
summary(D_model1)
plot(D_model1)
AIC(D_model1) #490.5862

diesel_nooutlier <- diesel_complete[-c(60,92),]

D_model1_nooutlier <- lm(Total.Exposure.Conc ~ Population + URBAN_RURAL, 
               diesel_nooutlier)
summary(D_model1_nooutlier)
plot(D_model1_nooutlier)
AIC(D_model1_nooutlier) #297.3357

D_model2 <- lm(Total.Exposure.Conc ~ `Population Density (Persons Per SqMile)` +
                 URBAN_RURAL, diesel_nooutlier)
summary(D_model2)
plot(D_model2)
AIC(D_model2) #410.9054

D_model2_nooutlier <- diesel_nooutlier[-c(31,42,64),]

D_model3 <- lm(Total.Exposure.Conc ~ `Population Density (Persons Per SqMile)` +
                 URBAN_RURAL, D_model2_nooutlier)
summary(D_model3)
plot(D_model3)
AIC(D_model3) #321.977

diesel_complete <- diesel_complete %>%
  mutate(log_conc = log10(Total.Exposure.Conc))

#log concentrations in diesel complete (new column)
#instead of total concentratuion, have log concentration in the model 

D_model_log <- lm(log_conc ~ Population + URBAN_RURAL, 
               diesel_complete)
summary(D_model_log)
plot(D_model_log)
AIC(D_model_log)

log_model2 <- lm(log_conc ~ Population + URBAN_RURAL, 
                diesel_complete)
summary(log_model2)
plot(log_model2)

diesel_nooutlier2 <- diesel_complete[-c(60,92),]
log_nooutliers <- diesel_nooutlier2[-c(32,41),]

log_model3 <- lm(log_conc ~ Population + URBAN_RURAL, 
                log_nooutliers)
summary(log_model3)
plot(log_model3)

log_nooutliers2 <- log_nooutliers[-c(26,33),]

log_model4 <- lm(log_conc ~ Population + URBAN_RURAL, 
                 log_nooutliers2)
summary(log_model4)
plot(log_model4)


```

