---
title: "JensenNarleskiSaraf_proj"
author: "Samantha Jensen, Jaclyn Narleski, Akanksha Saraf"
date: "2025-03-12"
output: html_document
---
Is there a significant difference between exposures for pollutants based on activity?
recreational, commercial, industrial, agricultural comparing specific pollutants (ANOVA)
comparing the NC air quality levels for exposure to the real levels that we are seeing.

next steps: 
- picking pollutants (2) (formaldehyde, diesel PM, maybe another)
- sorting columns into categories 
- find NC air quality standards for selected pollutants 


# Setup

```{r}

# Load packages
library(tidyverse)
library(here)
library(knitr)

# Load data
exposure <- read.csv("ExposureConcentrationsRaw.csv")

```
# Methods

## Data Source(s)

## Data Tidying

The raw dataset for exposure concentrations of air toxics in North Carolina contains the following information for each measurement of exposure: EPA region, county, FIPS code, FIPS tract, population, pollutant name, total exposure concentration, and several different sources that contributed to that total exposure concentration. The EPA divides these sources into different categories, as displayed below in Figure 1: 

Figure 1: National Air Toxics Assessment (NATA) Emission Sources

| On Road                       | Non-Road                           | Non-Point Stationary                       | Other                 |
|-------------------------------|------------------------------------|--------------------------------------------------------------------|
| Light Duty Gas On-Network     | Recreational & Pleasure Craft      | Industrial                                 | Fires                 |
| Light Duty Diesel On-Network  | Construction                       | Oil & Gas                                  | Biogenics             |
| Light Duty Gas Off-Network    | Commercial Lawn & Garden           | Commercial Cooking                         | Secondary Formation   |
| Light Duty Diesel Off-Network | Residential Lawn & Garden          | Solvents and Coatings                      | Background            |
| Heavy Duty Gas On-Network     | Agricultural Equipment             | Bulk Terminal Storage & Transfer (stage 1) |                       |                       
| Heavy Duty Diesel On-Network  | Commercial Equipment               | Fuel Combustion (except residential wood)  |                       |                       
| Heavy Duty Gas Off-Network    | Commercial Marine Vessel C1/C2     | Waste Disposal                             |                       |                       
| Heavy Duty Diesel Off-Network | Commercial Marine Vessel C3        | Agricultural Livestock                     |                       |                       
| Heavy Duty Diesel Hoteling    | Commercial Marine Vessels Underway | Miscellaneous Nonindustrial                |                       |                       
| Refueling                     | Locomotives                        |                                            |                       |                       
|                               | Airports (point source)            |                                            |                       |                     
|                               | Railroads (point source)           |                                            |                       |                       
|                               | All Other Non-Road Equipment       |                                            |                       |                       

"On Road" and "Non-Road" refer to mobile sources that emit air pollutants. "On Road" references mobile sources on the U.S. highway system that transport either passengers or freight. This includes passenger cars and light trucks (i.e. "light duty vehicles"); semi-trucks, RVs, delivery trucks, and buses (i.e. "heavy duty vehicles"); and each type of vehicle is separated by what type of fuel they use, gas or diesel. These types of vehicles are further categorized to "on-netork," or in motion, and "off-netork," or idle, as well as "refueling," or any vapor/spillage from fueling refills, or "hoteling," which is specific to heavy duty vehicles idling.

"Non-Road" references mobile sources that aren't on the U.S. highway system. This can range from recreational vehicles like snowcraft, boats, ATVs, dirt bikes to industrial/agricultural equipment and other forms of transportation like airports, railines ("railroads" and "locomotives"), and marine vessels. Marine vessels are categorized by engine size, with C1 and C2 considered smaller and C3 considered larger, as well as by "underway", or not anchored.

"Non-Point Stationary" refers to sources that aren't mobile or confined to single location. "Other" groups any other sources, including living organisms ("biogenics"), pollutants that form in the atmosphere from chemical reactions ("secondary"), and naturally-occurring emissions from human activity ("background").

```{r}

# Seperating out potential VOCs that are considered carcinogens
carcinogens <- exposure %>%
  filter(Pollutant.Name %in% c("FORMALDEHYDE", "BENZENE", "TOULENE", 
                               "CARBON TETRACHLORIDE", "CARBON DISULFIDE", 
                               "1,3-BUTADIENE", "1,4-DICHLOROBENZENE"),
         FIPS == "37063")

carcinogens_summary <- carcinogens %>%
  group_by(Pollutant.Name) %>%
  summarise(average_exposure = mean(Total.Exposure.Conc, na.rm = TRUE)) %>%
  ungroup()

```

# Initial Figures 

```{r}

exposure_fig1 <- ggplot() +
  geom_jitter(data = carcinogens,
              aes(x = 1, y = Total.Exposure.Conc, color = "maroon2"),
              alpha = 0.7) +
  geom_point(data = carcinogens_summary,
             aes(x = 1, y = average_exposure), 
             size = 4,
             color = "darkred") +
  facet_wrap(~Pollutant.Name, scales = "free_y") +
  labs(title = "Air Exposure Concentrations of Carcinogenic VOCs in Durham, NC",
       y = "Total Exposure Concentration (mg/m3)",
       x = NULL,
       caption = "The above graphs display the varying exposure concentrations 
       of recorded volatilie organic compounds (VOCs) in the air during 2014 in 
       Durham, North Carolina. These specific VOCs are 
       known to potentially increase the risk of cancer, making them 
       carcinogens.") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.caption = element_text(hjust = 0.5),
        plot.title = element_text(hjust = 0.5, face = "bold", color = "darkred"),
        legend.position = "none")

exposure_fig1
```

```{r}
carcinogens_activity_summary <- carcinogens %>%
  group_by(Pollutant.Name) %>% 
  summarise(across(c(PT.StationaryPoint.Exposure.Conc:NP.AgricultureLivestock.Conc), mean))

carcinogens_long <- pivot_longer(carcinogens_activity_summary, cols = c(PT.StationaryPoint.Exposure.Conc:NP.AgricultureLivestock.Conc),  names_to = "Variable", values_to = "Value")
carcinogens_long <- carcinogens_long %>% 
  filter(Variable == "NR.CommercialLawnGarden.Exposure.Conc"| Variable == 
           "NR.ResidentialLawnGarden.Exposure.Conc"| Variable == 
           "NR.Construction.Exposure.Conc"| Variable ==
           "NR.CommercialEquipment.Exposure.Conc")
carcinogens_long <- carcinogens_long %>% 
  filter(Pollutant.Name == "1,3-BUTADIENE"| Pollutant.Name == "BENZENE"| Pollutant.Name =="FORMALDEHYDE")

carcinogens_activity_summary <- carcinogens_activity_summary %>% 
  filter(Pollutant.Name == "1,3-BUTADIENE"| Pollutant.Name == "BENZENE"| Pollutant.Name =="FORMALDEHYDE")


figure3 <- ggplot(carcinogens_long, aes(x = Pollutant.Name, y = Value)) + 
  geom_point() + 
  facet_wrap(~Variable) + 
  labs(x = "Pollutant Name", 
       y = "Exposure Concentration (mg/m3)", 
       title = "Non-Zero Carcinogen Exposure Levels", 
       caption = "This figure depicts the mean carcinogen exposure levels in Durham County from certain activities.")
figure3

```

